Automating Political Bias Prediction

### 2016-06-13 manuscript/poltext.tex
Changed style from ICML to the ACL style file that the organisers sent around. Shortened text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2015 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2015,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

\usepackage{natbib}
\usepackage{amsmath,amsfonts,amscd,amssymb}
\usepackage{dsfont}
\renewcommand{\vec}[1]{\mathbf{#1}}
\usepackage{hyperref}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Corr}{Corr}
\newcommand{\R}{\mathds{R}}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{pbox}

@@

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2015}

% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Proceedings of the...''
%\usepackage[accepted]{icml2015}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Automating Political Bias Prediction}

\begin{document}

\twocolumn[
\icmltitle{Automating Political Bias Prediction}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2015
% package.
\icmlauthor{Felix Biessmann}{felix.biessmann@gmail.com}
\icmladdress{}
\icmlauthor{Pola Lehmann}{pola.lehmann@wzb.eu}
\icmladdress{Wissenschaftszentrum Berlin}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{}

\vskip 0.3in
]

@@

One area which produces large amounts of text which are of high interest for political scientists is modern media. Standard newspapers, but also new forms as twitter, facebook or blogs produce an ever growing amount of text influencing political debate. These texts can be of interest to political scientist in various analyses. The question we will focus on in this analysis is the potential political bias that can be detected in these sources. In many cases it is obvious which political bias an author has. In other cases some expertise is required to judge the political bias of a text. Keeping an unbiased view on what media report on requires to understand the political bias of texts. Different newspaper outlets are, for example, often ascribed as favoring different positions on the political left-right axis. Data from the European Media Systems Survey show that experts clearly associate different media outlets with specific parties (http://www.mediasystemsineurope.org). However such expert surveys on the political bias of newspapers can normally only give us a very general tendency of a newspaper, they do not distinguish between different journalists or different sections. Assistive technology can help in this context to try and obtain a more unbiased sample of information. It gives us the possibility to get a more nuanced view of the spectrum of political positions favored in important news outlets in a country. In this paper we will test whether it is possible to automatically detect political bias in text and if such a model can be used on different kind of texts. The aim of this study is to provide some empirical evidence indicating that leveraging open data sources automated political bias prediction is possible with above chance accuracy. We will test our model on both in-domain and out-of-domain data. Prediction on out-of-domain data is less precise as on in-domain data, but still generating satisfying results.

@@

@@

\subsection{Data}
Annotated political text data was obtained from two sources: a) the plenary debates held in the German parliament ({\em Bundestag}) and b) all manifesto texts of parties winning seats in the election to the German parliament in the current 18th and the last, 17th, legislative period.

@@

\paragraph{Parliament discussion data} Parliament texts are annotated with the respective party label, which we take here as a proxy for political bias. The protocols of plenary debates are available through the website of the German Bundestag\footnote{\url{https://www.bundestag.de/protokolle}}; an open source API was used to query the data in a cleaned and structured format\footnote{\url{https://github.com/bundestag}}. In total 22784 speeches were extracted for the 17th legislative period and 11317 speeches for the 18th period, queried until March 2016.

\paragraph{Party manifesto data}
The party manifesto text was taken from the Manifesto Corpus (add citation). The Corpus is accessible through an publicly available API published by the {\em Manifesto Project} \cite{manifesto}. The data released in this project mainly comprises the complete manifestos of all parties that have won seats at a national election. Each quasi-sentence\footnote{A quasi-sentence has the length of an argument. It is never longer than one sentence.} is annotated with one of 56 policy issue categories. Examples for the policy categories are {\em welfare state expansion, welfare state limitation, democracy, equality}; for a complete list and detailed explanations on how the annotators were instructed see \cite{leftright}. In total the 9 manifestos from the 17th and 18th legislative period have been divided into 29451 quasi-sentences. Each observation has been added with two labels: the party affiliation and the manually assigned policy issue aimed at in each specific statement.

@@

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
First each data set was segmented into semantic units; in the case of the party manifesto data the semantic units were the quasi-sentences associated with one of the 56 policy issues, in the case of parliament discussions this were the speeches. Parliament speeches were often interrupted; in this case each uninterrupted part of a speech was considered a semantic unit. Strings of each semantic unit were tokenised and transformed into bag-of-word vectors as implemented in scikit-learn \cite{scikit-learn}. The general idea of bag-of-words vectors is to simply count occurrences of words (or word sequences, also called {\em n-grams}) for each data point. A data point is usually a document, here it is the semantic units of parliament speeches and manifesto sentences, respectively. The text of each semantic unit is transformed into a vector $\vec{x}\in\mathds{R}^d$ where $d$ is the size of the dictionary; the $w$th entry of $\vec{x}$ contains the (normalized) count of the $w$th word (or sequence of words) in our dictionary. Several options for vectorizing the speeches were tried, including term-frequency-inverse-document-frequency normalisation, n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

@@

Three different classification problems were considered:
\begin{enumerate}
\item {\bf Classification of party affiliation} (five class / four class problem)
\item {\bf Classification of government membership} (binary problem)
\item {\bf Classification of policy issues} (56 class problem)
\end{enumerate}

Party affiliation is a five class problem for the 17th legislation period, and a four class problem for the 18th legislation period. The policy issue classification is based on the categories of the Manifesto Project, see \autoref{sec:data} and \cite{leftright}.
For both the party affiliation and the government membership prediction, classifiers were trained on the parliament speeches. For the third problem classifiers were trained only on the manifesto data for which policy issue labels were available.

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data; wherever possible the generalisation performance of the models was also evaluated on out-of domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set only: A cross-validation split was made to obtain train/test data for the grid search and for each setting of hyperparameters the entire pipeline was trained and evaluated -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Out-of-domain evaluation data were the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).
% Falls wir die Medien nochreinnehmen, muss das bei out-of-domain auch noch erwähnt werden.

@@

\subsection{Analysis of bag-of-words features}\label{sec:correlations_methods}
While interpretability of linear models is often propagated as one of their main advantages, doing so naively without modelling the noise covariances can lead to wrong conclusions, see e.g. \cite{Zien2009, Haufe2013}; interpreting coefficients of linear models (independent of the regularizer used) implicitly assumes uncorrelated features; this assumption is violated by the text data used in this study. Thus direct interpretation of the model coefficients $\vec{W}$ is problematic. In order to allow for better interpretation of the predictions and to assess which features are discriminative correlation coefficients between each word and the party affiliation label were computed. The words corresponding to the top positive and negative correlations are shown in \autoref{sec:word_party_correlations}.

@@

When predicting party affiliation on text data from the same domain that was used for training the model, average precision and recall values of above 0.6 are obtained. These results are comparable to those of \cite{Hirst2014} who report a classification accuracy of 0.61 on a five class problem predicting party affiliation in the European parliament; the accuracy for the 17th Bundestag is 0.63, results of the 18th Bundestag are difficult to compare as the number of parties is four and the legislation period is not finished yet.

But as the main purpose of this paper is to generate a model that can be used to predict political bias on new data, like media texts, we are interested in the question if the model also produces satisfying results on out-of-domain data. We are therefore also predicting party affiliation on party manifesto text. For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4 \autoref{tab:results_out-of-domain}. This drop in out of domain prediction accuracy is in line with previous findings \cite{Yu2008}.
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due to the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, the manifesto data was aggregated into the previously described party domains, see \autoref{sec:data}. The topic level results are shown in \autoref{tab:results_topic} and demonstrate that when the texts to be classified are sufficiently long and the word count statistics are sufficiently dense the classification performance on out-of-domain data can - at least in some cases - achieve reliable precision and recall values close to 1.0. This increase is in line with previous findings on the influence of text length on political bias prediction accuracy \cite{Hirst2014}.

@@

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on out-of-domain data.  Predictions are done on the {\bf sentence level}.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 \\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 \\
     gruene   &    0.47    &  0.20   &   0.28    &  3747\\
      linke     &  0.30  &    0.47    &  0.37    &   1701\\
        spd     &  0.26  &    0.16   &   0.20    &   2278\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see \autoref{tab:results_17}, \autoref{tab:results_18}, \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases. Note that all manifesto topics of the green party in the 18th Bundestag are predicted to be from the parties of the governing coalition, CDU/CSU or SPD.}
\begin{center}
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf Party Manifestos} \\
& \multicolumn{4}{c}{\bf 17th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

@@

\subsection{Validation}
\paragraph{Differentiability of policy issues}
One reason why some statements are difficult to classify could be that some issues are more difficult to differentiate than others. To test this a separate suite of experiments was run to train and test the prediction performance of the text classifiers models described in \autoref{sec:model}. As there was no out-of-domain evaluation set available in this setting only the evaluation error on in-domain data is reported. Note however that in this experiment too the evaluation data was never seen by any model during training time.
In \autoref{tab:results_avg_political_view} results for the best and worst classes, in terms of predictability, are listed along with the average performance metrics on all classes.
Precision and recall values of close to 0.5 on average can be considered rather high considering the large number of labels. Results are not surprising, but are equal to those by human coders. \\

\begin{table}
\caption{
\label{tab:results_avg_political_view}
Classification of 56 political views, see \autoref{sec:data}.
}
\begin{center}
\footnotesize
\begin{tabular}{lcrrrr}
code & meaning  &      prec    &recall & N\\
\hline\hline
%
       501  & environmentalism + &      0.62   &   0.61 &      165\\
        202 &   democracy + &   0.58  &    0.55   &      122\\
        701    & labour +&  0.57  &    0.54   &   129\\
                201    &freedom/human rights +  & 0.58   &   0.54   &      159\\
         106   & peace + & 0.52&      0.57   &       21\\
\dots\\
        302     &centralism + & 0.25     & 0.20  &          10\\
        401    &  free enterprise + &0.20    &  0.19    &       52\\
        505    &welfare - &  0.13   &   0.14   &      14\\
        409    & keynesian demand +&  0.14  &      0.13   &      8\\
            0    & undefined &  0.09  &    0.12   &    17\\
            \hline
avg / total  &  &  0.47    &  0.46 &    2946\\
%
\end{tabular}
\end{center}
\end{table}

\paragraph{Policy change}
Another explanation for missclassification could lie in the fact that parties change their policy positions. In order to investigate this confusion matrices were extracted for the predictions on the out-of-domain evaluation data for sentence level predictions (see \autoref{tab:confusion}) as well as topic level predictions (see \autoref{tab:confusion_topic}). On the topic level results for the 17th legislative period are pretty good except for the SPD. In the 18th legislative period predictions are very bad for the Green party. One explanation for this missclassification could lie in policy changes. For example, the Green party has devoted large parts of its manifesto to renewable and against nuclear energy. After Fukushima the other parties adopted these positions. \\

% Die könnten wir vielleicht rausnehmen.
% Confusion matrix sentence level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion} {\bf Confusion matrices (sentence level)} for predictions on evaluation data (party manifestos); classifiers were trained on parliament speeches for the 17th legislative period; the most prominent effect is the high likelihood for a party to be taken as the strongest, governing party, cdu/csu. This can be interpreted as a change in policies of the conservative party cdu/csu towards the policies of the green party.}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}}& &cducsu &1186 &289& 178& 198& 179\\
&&fdp &882& 658& 236& 329& 214\\
&&gruene &1174& 404& 764& 941& 464\\
&&linke &388& 92& 214& 806& 201\\
&&spd &999& 268& 240& 398& 373\\
\end{tabular}
\end{table}

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Confusion matrices (topic level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right).}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}} &&cducsu &7& 0& 0& 0& 0\\
&&fdp&0& 7& 0& 0& 0\\
&&gruene&0& 0& 6& 0& 1\\
&&linke&0& 0& 0& 7& 0\\
&&spd&4& 0& 0& 0& 4\\
\end{tabular}
\end{table}

@@

%
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\caption{
\label{tab:results_binary_18}
Classification performance on the binary prediction problem in the 18th legislative period, categorizing speeches into government (SDP/CDU/CSU) and opposition (Linke, Gr\"une).
}
\begin{center}
\begin{tabular}{lcccccccc}
& \multicolumn{4}{c}{\bf Held-out parliament speeches} & \multicolumn{4}{c}{\bf Party Manifestos}\\
    &         precision    &recall &  f1-score  & N    &         precision    &recall &  f1-score  & N\\
\hline \hline
government   &    0.88  &    0.95    &  0.92   &    786   &0.52   &   0.66 &     0.58  &    5972\\
 opposition    &   0.86    &  0.71   &   0.78   &    346 & 0.69  &    0.56    &  0.62   &   8229\\
\hline
avg / total    &   0.88    &  0.88     & 0.87  &    1132 &  0.62   &   0.60    &  0.60 &    14201\\
%
\end{tabular}
\end{center}
\end{table*}

@@

\subsection{Predicting government status}\label{sec:sentiment_result}
Next to the party affiliation labels also government membership labels were used to train models that predict whether or not a text is from a party that belonged to a governing coalition of the Bundestag. In \autoref{tab:results_binary_17} the results are shown for the 17th legislative period. While the in-domain evaluation precision and recall values reach values close to 0.9, the out-of-domain evaluation drops again to values between 0.6 and 0.7. This is in line with the results on binary classification of political bias in the Canadian parliament \cite{Yu2008}. The authors report classification accuracies between 0.8 and 0.87, the accuracy in the 17th Bundestag was 0.85. While topic-level predictions were not performed in this binary setting, the party affiliation results in \autoref{tab:results_topic} suggest that a similar increase in out-of-domain prediction accuracy could be achieved when aggregating texts to longer segments. \\

@@

The results show an interesting relationship between political power and sentiment. Political power was evaluated in terms of membership of the government. Correlating these indicators of political power with the mean sentiment of a party shows a strong positive correlation between speech sentiment and political power. This pattern is evident from the data in \autoref{fig:party_sentiments} and in \autoref{tab:sentiments}: In the current Bundestag, government membership correlates with positive sentiment with a correlation coefficient of 0.98 and the number of seats correlates with 0.89.

Note that there is one party, the social democrats (SPD), which has many seats and switched from opposition to government with the 18th Bundestag: With its participation in the government the average sentiment of this party switched sign from negative to positive, suggesting that positive sentiment is a strong indicator of government membership.

>>>>

%
% File poltext2016.tex
%
% Contact: jan.snajder@fer.hr or dsirinic@gmail.com
%%
%% Based on the style files for ACL-2015, which were, in turn,
%% Based on the style files for ACL-2014, which were, in turn,
%% Based on the style files for ACL-2013, which were, in turn,
%% Based on the style files for ACL-2012, which were, in turn,
%% based on the style files for ACL-2011, which were, in turn, 
%% based on the style files for ACL-2010, which were, in turn, 
%% based on the style files for ACL-IJCNLP-2009, which were, in turn,
%% based on the style files for EACL-2009 and IJCNLP-2008...

%% Based on the style files for EACL 2006 by 
%%e.agirre@ehu.es or Sergi.Balari@uab.es
%% and that of ACL 08 by Joakim Nivre and Noah Smith

\documentclass[11pt]{article}
\usepackage{poltext2016}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

\usepackage{natbib}
\usepackage{amsmath,amsfonts,amscd,amssymb}
\usepackage{dsfont}
\renewcommand{\vec}[1]{\mathbf{#1}}
\usepackage{hyperref}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Corr}{Corr}
\newcommand{\R}{\mathds{R}}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{color}

@@

\newcommand{\pola}[1]{\textcolor{blue}{#1}}
\newcommand{\felix}[1]{\textcolor{green}{#1}}

%\setlength\titlebox{5cm}

% You can expand the titlebox if you need extra space
% to show all the authors. Please do not make the titlebox
% smaller than 5cm (the original size); we will check this
% in the camera-ready version and ask you to change it back.

\title{Automating Political Bias Prediction}

\author{Felix Biessmann\\
  {\tt felix.biessmann@gmail.com} \\\And
 Pola Lehmann \\
  Wissenschaftszentrum Berlin \\
  {\tt pola.lehmann@wzb.eu} \\}

\date{}

\begin{document}
\maketitle

@@

One area which produces large amounts of text which are of high interest for political scientists is modern media. Standard newspapers, but also new forms as twitter, facebook or blogs produce an ever growing amount of text influencing political debate. These texts can be of interest to political scientist in various analyses. The question we will focus on in this analysis is the potential political bias that can be detected in these sources. In many cases it is obvious which political bias an author has. In other cases some expertise is required to judge the political bias of a text. Keeping an unbiased view on what media report on requires to understand the political bias of texts. Different newspaper outlets are, for example, often ascribed as favoring different positions on the political left-right axis. Data from the European Media Systems Survey show that experts clearly associate different media outlets with specific parties (http://www.mediasystemsineurope.org). However such expert surveys on the political bias of newspapers can normally only give us a very general tendency of a newspaper, they do not distinguish between different journalists or different sections. Assistive technology can help in this context to try and obtain a more unbiased sample of information. It gives us the possibility to get a more nuanced view of the spectrum of political positions favored in important news outlets in a country. In this paper we will test whether it is possible to automatically detect political bias in text and if such a model can be used on different kind of texts. The aim of this study is to provide some empirical evidence indicating that leveraging open data sources automated political bias prediction is possible with above chance accuracy. We will test our model on both in-domain and out-of-domain data; the different train and test data sets are summarized in \autoref{tab:classification_settings}. The classification tasks we considered as proxies for political bias prediction are

\begin{enumerate}
\item {\bf Classification of party affiliation} (five class problem)
\item {\bf Classification of government membership} (binary problem)
\item {\bf Classification of policy issues} (56 class problem)
\end{enumerate}

@@

% Confusion matrix topic level
\begin{table}[t]\label{tab:classification_settings}
\footnotesize
\begin{tabular}{ccc}
Training Set & \multicolumn{2}{c}{Test set} \\
& In-domain&Out-of-domain\\
\hline
\hline
Parliament ($90\%$) & Parliament ($10\%$)& Manifestos\\
Manifestos ($90\%$)& Manifestos ($10\%$) & -\\
\end{tabular}
\end{table}

@@

\subsection{Data}
Annotated political text data was obtained from two sources: a) the plenary debates held in the German parliament ({\em Bundestag}) and b) all manifesto texts of parties winning seats in the election to the German parliament in the the last, 17th, legislative period.

@@

\paragraph{Parliament discussion data} Parliament texts are annotated with the respective party label, which we take here as a proxy for political bias. The protocols of plenary debates are available through the website of the German Bundestag\footnote{\url{https://www.bundestag.de/protokolle}}; an open source API was used to query the data in a cleaned and structured format\footnote{\url{https://github.com/bundestag}}. In total 22784 speeches were extracted for the 17th legislative period.

\paragraph{Party manifesto data}
The party manifesto text was taken from the Manifesto Corpus  \cite{manifesto}. The data released in this project mainly comprises the complete manifestos of all parties that have won seats at a national election. Each quasi-sentence\footnote{A quasi-sentence has the length of an argument. It is never longer than one sentence.} is annotated with one of 56 policy issue categories. Examples for the policy categories are {\em welfare state expansion, welfare state limitation, democracy, equality}; for a complete list and detailed explanations on how the annotators were instructed see \cite{leftright}. Each quasi-sentence has  two types of labels: the party affiliation and the manually assigned policy issue aimed at in each specific statement.

@@

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
All text data was segmented into semantic units, quasi-sentences for manifesto data and (uninterrupted) speech parts for parliament speech data. Strings of each semantic unit were tokenised and transformed into bag-of-word (BOW) vectors as implemented in scikit-learn \cite{scikit-learn}. Several options for BOW vectorizations were tried, including term-frequency-inverse-document-frequency normalisation, n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

@@

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Wherever possible the generalisation performance of the models was also evaluated on out-of domain texts, for which we used the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).
% Falls wir die Medien nochreinnehmen, muss das bei out-of-domain auch noch erwähnt werden.

@@

\subsection{Interpreting bag-of-words models}\label{sec:correlations_methods}
Interpreting coefficients of linear models (independent of the regularizer used) implicitly assumes uncorrelated features; this assumption is violated by the text data used in this study. Thus direct interpretation of the model coefficients $\vec{W}$ is problematic, see also \cite{Zien2009, Haufe2013}. In order to allow for better interpretation of the predictions and to assess which features are discriminative correlation coefficients between each word and the party affiliation label were computed. The words corresponding to the top positive and negative correlations are shown in \autoref{sec:word_party_correlations}.

@@

When predicting party affiliation on text data from the same domain that was used for training the model, average precision and recall values of above 0.6 are obtained. These results are comparable to those of \cite{Hirst2014} who report a classification accuracy of 0.61 on a five class problem predicting party affiliation in the European parliament.
For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4 \autoref{tab:results_out-of-domain}. This drop in out of domain prediction accuracy is in line with previous findings \cite{Yu2008}.
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due to the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, the manifesto data was aggregated into the previously described party domains. The topic level results are shown in \autoref{tab:results_topic} and demonstrate that when the texts to be classified are sufficiently long and the word count statistics are sufficiently dense the classification performance on out-of-domain data can - at least in some cases - achieve reliable precision and recall values close to 1.0. This increase is in line with previous findings on the influence of text length on political bias prediction accuracy \cite{Hirst2014}.

@@

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on out-of-domain data.  Predictions are done on the {\bf sentence level}.\label{tab:confusion}
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 \\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 \\
     gruene   &    0.47    &  0.20   &   0.28    &  3747\\
      linke     &  0.30  &    0.47    &  0.37    &   1701\\
        spd     &  0.26  &    0.16   &   0.20    &   2278\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see  also \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases.}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

@@

\subsection{Validation}
\paragraph{Differentiability of policy issues}
One reason why some statements are difficult to classify could be that some issues are more difficult to differentiate than others. To test this a separate suite of experiments was run to train and test the prediction performance of the text classifiers models described in \autoref{sec:model}. As there was no out-of-domain evaluation set available in this setting only the evaluation error on in-domain data is reported. Note however that in this experiment too the evaluation data was never seen by any model during training time.
In \autoref{tab:results_avg_political_view} results for the best and worst classes, in terms of predictability, are listed along with the average performance metrics on all classes.
Precision and recall values of close to 0.5 on average can be considered rather high considering the large number of labels. \felix{Results are not surprising, but are equal to those by human coders.} \\

\begin{table}
\caption{
\label{tab:results_avg_political_view}
Classification of 56 political views, see \autoref{sec:data}.
}
\begin{center}
\footnotesize
\begin{tabular}{lcrrrr}
code & meaning  &      prec    &recall & N\\
\hline\hline
%
       501  & environmentalism + &      0.62   &   0.61 &      165\\
        202 &   democracy + &   0.58  &    0.55   &      122\\
        701    & labour +&  0.57  &    0.54   &   129\\
                201    &human rights +  & 0.58   &   0.54   &      159\\
         106   & peace + & 0.52&      0.57   &       21\\
\dots\\
        302     &centralism + & 0.25     & 0.20  &          10\\
        401    &  free enterprise + &0.20    &  0.19    &       52\\
        505    &welfare - &  0.13   &   0.14   &      14\\
        409    & keyn. demand +&  0.14  &      0.13   &      8\\
            0    & undefined &  0.09  &    0.12   &    17\\
            \hline
 total  &  &  0.47    &  0.46 &    2946\\
%
\end{tabular}
\end{center}
\end{table}

\paragraph{Policy change}
Another explanation for missclassification could lie in the fact that parties change their policy positions. In order to investigate this confusion matrices were extracted for the predictions on the out-of-domain evaluation data for sentence level predictions (see \autoref{tab:confusion}) as well as topic level predictions (see \autoref{tab:confusion_topic}). On the topic level results for the 17th legislative period are pretty good except for the SPD, which the classifier often predicts as CDU/CSU. One explanation for this missclassification could lie in policy changes, due to e.g. the coalition contract that forces the SPD to eventually accept compromises with the other governing parties, CDU/CSU. \\

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Topic level confusion matrices} on evaluation data (manifesto texts), classifiers were trained on speeches of 17th Bundestag.}
\vspace{0.5em}
\footnotesize
\begin{tabular}{lcccccc}
&& \multicolumn{5}{c}{\bf Predicted}\\
&& cducsu & fdp& gruene& linke& spd\\
\hline
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{2cm}{\centering {\bf True}}}} &cducsu &7& 0& 0& 0& 0\\
&fdp&0& 7& 0& 0& 0\\
&gruene&0& 0& 6& 0& 1\\
&linke&0& 0& 0& 7& 0\\
&spd&4& 0& 0& 0& 4\\
\end{tabular}
\end{table}

@@

%
\end{tabular}
\end{center}
\end{table*}
%%
%%\begin{table*}[t]
%%\caption{
%%\label{tab:results_binary_18}
%%Classification performance on the binary prediction problem in the 18th legislative period, categorizing speeches into government (SDP/CDU/CSU) and opposition (Linke, Gr\"une).
%%}
%%\begin{center}
%%\begin{tabular}{lcccccccc}
%%& \multicolumn{4}{c}{\bf Held-out parliament speeches} & \multicolumn{4}{c}{\bf Party Manifestos}\\
%%    &         precision    &recall &  f1-score  & N    &         precision    &recall &  f1-score  & N\\
%%\hline \hline
%%government   &    0.88  &    0.95    &  0.92   &    786   &0.52   &   0.66 &     0.58  &    5972\\
%% opposition    &   0.86    &  0.71   &   0.78   &    346 & 0.69  &    0.56    &  0.62   &   8229\\
%%\hline
%%avg / total    &   0.88    &  0.88     & 0.87  &    1132 &  0.62   &   0.60    &  0.60 &    14201\\
%%%
%%\end{tabular}
%%\end{center}
%%\end{table*}

@@

\subsection{Predicting government status}\label{sec:sentiment_result}
Next to the party affiliation labels also government membership labels were used to train models that predict whether or not a text is from a party that belonged to a governing coalition of the Bundestag. In \autoref{tab:results_binary_17} the results are shown for the 17th legislative period. While the in-domain evaluation precision and recall values reach values close to 0.9, the out-of-domain evaluation drops again to values between 0.6 and 0.7. This is in line with the results on binary classification of political bias in the Canadian parliament \cite{Yu2008}. The authors report classification accuracies between 0.80 and 0.87, the accuracy in the 17th Bundestag was 0.85. While topic-level predictions were not performed in this binary setting, the party affiliation results in \autoref{tab:results_topic} suggest that a similar increase in out-of-domain prediction accuracy could be achieved when aggregating texts to longer segments. \\

@@

The results show an interesting relationship between political power and sentiment. Political power was evaluated in terms of membership of the government. Correlating these indicators of political power with the mean sentiment of a party shows a strong positive correlation between speech sentiment and political power. This pattern is evident from the data in 
%\autoref{fig:party_sentiments} and 
in \autoref{tab:sentiments}: In the current Bundestag, government membership correlates with positive sentiment with a correlation coefficient of 0.98 and the number of seats correlates with 0.89.
